---
title: 'San Francisco Multifamily Real Estate Development'
author: 'Dean Boerner/ECON 631'
date: "5/12/2022"
output: ioslides_presentation
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width=7, fig.height=4)


library(tidyverse)
library(sf)
library("RSocrata")
library(rgdal)
library(viridis)
library(plotly)
library(ggrepel)  

#making the map theme from the textbook
theme_map <- function(base_size=9, base_family="") { require(grid)
theme_bw(base_size=base_size, base_family=base_family) %+replace% theme(axis.line=element_blank(),
axis.text=element_blank(), 
axis.ticks=element_blank(), 
axis.title=element_blank(), 
panel.background=element_blank(), 
panel.border=element_blank(), 
panel.grid=element_blank(), 
panel.spacing=unit(0, "lines"), 
plot.background=element_blank(), 
legend.justification = c(0,0), 
legend.position = c(0,0)) 
}

#reading in cleaned data
mf_raw <- read.csv('mf.csv', stringsAsFactors = FALSE)
mf_raw <- as_tibble(mf_raw)  

#converting to an sf object, where longitudes are x-coordinates, latitudes y-.
mf_sf <- st_as_sf(x=mf_raw, coords = c("Longitude", "Latitude"),
    agr = "constant",
    crs = 7131,
    stringsAsFactors = FALSE,
    remove = TRUE
    )

#Viz 1
url <- 'https://data.sfgov.org/resource/xfcw-9evu.geojson'
geo_object <- read_sf(url)
st_crs(geo_object) <- 7131

mf_joined <- st_join(mf_sf, geo_object, join=st_within)

#Vizzes 2, 3, 4
url2 <- 'https://data.sfgov.org/resource/keex-zmn4.geojson'
geo_object2 <- read_sf(url2)
st_crs(geo_object2) <- 7131
geo_object2 <- st_transform(geo_object2, 7131)

mf_supervisor_joined <- st_join(mf_sf, geo_object2, join=st_within)
```
## Introduction

- Task: Find where multifamily projects are being proposed and approved around the city of San Francisco  

- Motivation: Narrowing down number of neighborhoods to choose from in crafting a West Coast investment, development strategy 

- The Data: construction, entitlement activity on all ongoing project plans – posted quarterly by the city  
  - Cleaning involved filtering to projects with 2 or more residential units and to those having been updated in given quarterly reports.  
  
- In short: We can look at total activity of different kinds (e.g. brand-new filings, new approvals, new construction).

## Net Multifamily Units Starting Construction In S.F. Since 2015 
```{r Figure 1, fig.cap="Data Source: DataSF. Counting construction starts."}
#grouping for total number of filings by neighborhood
mf_sf_summed <- group_by(as_tibble(mf_joined), nhood) %>% 
  summarize('Number_of_Filings' = n())

#adding missing neighborhoods
other_neighs <- tibble(nhood = c("Golden Gate Park", "McLaren Park", "Seacliff", "Presidio", "Lincoln Park"), 
                       Number_of_Filings = c(0, 0, 0, 0, 0))
mf_sf_summed <- add_row(mf_sf_summed, other_neighs)

#joining the map data to total number of filings on neighborhood
mf_sf_summed_joined <- left_join(geo_object, mf_sf_summed)

#checking formatting of these specific neighborhood names and making a new column in the process
mf_sf_summed_joined$Neighborhood <- ifelse(mf_sf_summed_joined$nhood == 'Golden Gate Park', 'Golden Gate Park', 
                                           ifelse(mf_sf_summed_joined$nhood == 'Lincoln Park', 'Lincoln Park', 
                                                  ifelse(mf_sf_summed_joined$nhood == 'Seacliff', 'Seacliff', 
                                                         ifelse(mf_sf_summed_joined$nhood == 'Presidio', 'Presidio', 
                                                                ifelse(mf_sf_summed_joined$nhood == 'McLaren Park', 'McLaren Park', 
                                                                       mf_sf_summed_joined$nhood)))))

#adding progress columns to the original spatially joined dataframe
mf_joined$Progress <- ifelse(mf_joined$Status %in% c('BP Filed', 'BP FIled', 'BP FILED', 'PL Filed', 'PL FILED'), 'Filing', 
                             ifelse(mf_joined$Status %in% c('BP APPROVED', 'BP ISSUED', 'Construction', 'CONSTRUCTION', 
                                                            'PL Approved', 'PL APPROVED'), 'Approval', 'Neither'))
mf_joined$Planning_Filing <- ifelse(mf_joined$Status %in% c('PL Filed', 'PL FILED'), 'Planning Filing', 'Not')
mf_joined$Construction <- ifelse(mf_joined$Status %in% c('Construction', 'CONSTRUCTION'), 'Construction', 'Not')

#filtering to only new filings, grouping by neighborhood, counting new filings, summming net units by neighborhood for those new filings, 
#and converting to tibble for join with map data joined data.
mf_other_group <- mf_joined %>% 
  filter(New_Filing == 1 & Construction == 'Construction') %>% 
  group_by(nhood) %>% 
  summarize("New_Filings" = n(), "Multifamily_Units_Started" = sum(Net_Units)) %>% 
  as_tibble()

#removing superfluous geometry column
mf_other_group <- select(mf_other_group, -geometry)

#joining new filing tibble to map data
full_other_join <- left_join(mf_sf_summed_joined, mf_other_group)
#for neighborhood labels
full_other_join$centroid <- st_centroid(full_other_join$geometry)
full_other_join$X <- st_coordinates(full_other_join$centroid)[, 1]
full_other_join$Y <- st_coordinates(full_other_join$centroid)[, 2]


gg_map_other_pt1 <- ggplot(full_other_join) + geom_sf(aes(fill=Multifamily_Units_Started, text=paste("<b>", Neighborhood, 
                                                                                                     ": <b>", Multifamily_Units_Started, 
                                                                                                     "multifamily units started since 2015")))
gg_map_other_pt2 <- gg_map_other_pt1 + geom_text(aes(x=X, y=Y, label=Neighborhood), size=1.5) + 
  scale_fill_fermenter(palette = "YIGn", direction=1, labels=scales::comma) + labs(fill = "Net Units Started") + theme_map() + 
  theme(legend.background = element_blank())

ggplotly(gg_map_other_pt2, tooltip="text") %>% 
style(hoverlabel=list(bgcolor = "white"), hoveron = "fill")
```
## Planning Filing, Approval Progress by Supervisor District Since 2015
```{r Figure 2, fig.cap="Data Source: DataSF. Counting new planning applications, approvals"}
#fig 2

#defining progress in supervisor-joined data
mf_supervisor_joined$Progress <- ifelse(mf_supervisor_joined$Status %in% c('BP Filed', 'BP FIled', 'BP FILED', 'PL Filed', 'PL FILED'), 'Filing', ifelse(mf_supervisor_joined$Status %in% c('BP APPROVED', 'BP ISSUED', 'Construction', 'CONSTRUCTION', 'PL Approved', 'PL APPROVED'), 'Approval', 'Neither'))
mf_supervisor_joined$Planning_Filing <- ifelse(mf_supervisor_joined$Status %in% c('PL Filed', 'PL FILED'), 'Planning Filing', 
                                               ifelse(mf_supervisor_joined$Status %in% c('PL Approved', 'PL APPROVED'), 
'Planning Approval', 'Neither'))

#getting data to adjust for supervisor district size
geo_object2_modified <- geo_object2
geo_object2_modified$some_distance_squared <- st_area(geo_object2_modified$geometry)
geo_object2_modified$rough_square_miles <- geo_object2_modified$some_distance_squared * 3685.89
geo_object2_modified <- as_tibble(select(geo_object2_modified, supdist, rough_square_miles))
geo_object2_modified <- select(geo_object2_modified, -geometry)
geo_object2_modified$rough_square_miles <- as.numeric(geo_object2_modified$rough_square_miles)

#***test
rel_supe_by_progress <- mf_supervisor_joined %>% 
  filter(New_Filing == 1 & Planning_Filing %in% c('Planning Filing', 'Planning Approval')) %>% 
  group_by(supdist, Planning_Filing) %>% 
  summarize("Number_of_Steps" = sum(New_Filing), "Net_Units" = sum(Net_Units)) %>% 
  as_tibble()

rel_supe_by_progress <- left_join(rel_supe_by_progress, geo_object2_modified)
rel_supe_by_progress$adjusted_number_of_steps <- (rel_supe_by_progress$Number_of_Steps) / (rel_supe_by_progress$rough_square_miles)
rel_supe_by_progress$adjusted_net_units <- (rel_supe_by_progress$Net_Units) / (rel_supe_by_progress$rough_square_miles)

rel_supe_by_progress$District <- c(1, 1, 10, 10, 11, 11, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 9, 9)

title_mod = "Districts 6, 9: Most Activity"
title_2_mod = "6, 3: Most Activity Per Square Mile"

#could use seperate geom_texts with seperate agged tibbles to add labels for each fill
gross_counts_update <- ggplot(subset(mf_supervisor_joined, New_Filing == 1 & Planning_Filing != 'Neither'), 
                aes(x=reorder(supervisor, New_Filing, sum), fill=Planning_Filing))
gs2_update <- gross_counts_update + geom_bar(alpha=0.8) + 
  scale_fill_manual(values=c("springgreen3", "gold1"), labels=c("Approvals", "Applications")) + labs(title=title_mod, x="District", y="Planning Filings", fill="Filing Type") + 
  coord_flip() + theme(legend.position = c(0.88, 0.82), 
                       legend.background = element_blank(),
                       legend.key.size = unit(.5, 'cm'), #change legend key size
                       legend.title = element_text(size=5), #change legend title font size
                       legend.text = element_text(size=5)) #change legend text font size)

#adjusting for area
adjusted_counts_mod <- ggplot(rel_supe_by_progress, aes(x=reorder(District, adjusted_number_of_steps, sum), y=adjusted_number_of_steps, fill=Planning_Filing))
adj2_mod <- adjusted_counts_mod + geom_bar(stat = "identity", alpha=0.8) + scale_fill_manual(values=c("springgreen3", "gold1")) + guides(fill=FALSE) + labs(title=title_2_mod, x="District", y="Planning Filings per Square Mile") + coord_flip()

cowplot::plot_grid(gs2_update, adj2_mod, ncol=2, align = "h")
```
## Approvals-To-Applications Ratios By Supervisor District Since 2015
```{r Figure 3, fig.cap="Data Source: DataSF. Dividing number of planning approvals by applications"}
supe_approvals_to_filings <- tibble(supdist=c('SUPERVISORIAL DISTRICT 1', 'SUPERVISORIAL DISTRICT 2', 'SUPERVISORIAL DISTRICT 3',
                                              'SUPERVISORIAL DISTRICT 4', 'SUPERVISORIAL DISTRICT 5', 'SUPERVISORIAL DISTRICT 6',
                                              'SUPERVISORIAL DISTRICT 7', 'SUPERVISORIAL DISTRICT 8', 'SUPERVISORIAL DISTRICT 9',
                                              'SUPERVISORIAL DISTRICT 10', 'SUPERVISORIAL DISTRICT 11'),
                                    approvals_to_filings=c((1/16), (6/31), (4/35), (1/6), (11/37), (21/70), (2/5), (4/22), (11/46),
                                                          (12/23), (1/10)),
                                    Approvals=c(1, 6, 4, 1, 11, 21, 2, 4, 11, 12, 1),
                                    Filings=c(16, 31, 35, 6, 37, 70, 5, 22, 46, 23, 10),
                                    Square_Miles=c(4.4, 5.2, 1.8, 3.8, 2.5, 3.7, 7.6, 3.4, 3.1, 8.3, 3.0))
supe_approvals_to_filings_joined <- left_join(supe_approvals_to_filings, geo_object2)
supe_approvals_to_filings_joined <- st_as_sf(x=supe_approvals_to_filings_joined, agr = "constant", crs = 7131, stringsAsFactors = FALSE)

supe_approvals_to_filings_joined$centroid <- st_centroid(supe_approvals_to_filings_joined$geometry)
supe_approvals_to_filings_joined$X <- st_coordinates(supe_approvals_to_filings_joined$centroid)[, 1]
supe_approvals_to_filings_joined$Y <- st_coordinates(supe_approvals_to_filings_joined$centroid)[, 2]

sup_plot1 <- ggplot(supe_approvals_to_filings_joined) + geom_sf(aes(fill=approvals_to_filings, 
                                                                    text=paste("District", "<b>", supervisor, "(", "<b>", Square_Miles, "sq. mi.): ", "<b>", Approvals, "approvals and", "<b>", Filings, "applications since 2015")))

sup_plot2 <- sup_plot1 + geom_text(aes(x=X, y=Y, label=supervisor)) + scale_fill_fermenter(palette = "YIGn", direction=1) + 
  labs(fill="Approvals/Applications") + theme_map() 
ggplotly(sup_plot2, tooltip="text") %>%
style(hoverlabel=list(bgcolor = "white"), hoveron = "fill")
```

## Conclusion

- The Bayview Hunters Point and Potrero Hill neighborhoods might offer the most attractive combination of higher growth trajectory and lower disapproval risk.  
- I recommend that the firm's current property selection efforts zero in on District 10 sites but that it hold off on any action until completing further research. Caveats include:  
  - Where is high vs. low saturation a flaw versus a feature?  
  - Where do high approvals relative to filings indicate approval tendencies rather than a dip in filings?  
  
- At the very least, high total and relative number of approvals in Bayview Hunters Point, Potrero Hill, and neighboring submarkets warrant serious consideration.

